Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-30 report/chapter01.Rnw
.

Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty. Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices. A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals. Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background. We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction. All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.

This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence. Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required. The following sections outline our contributions, the research questions we address, and the scope of the work.

\section{Contributions and research questions}

We make three contributions within a single, notation-coherent framework.
\begin{enumerate}
  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
\end{enumerate}

Two questions motivate the empirical study.
\begin{enumerate}
  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines? We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)? We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
\end{enumerate}

\section{Scope and positioning}

We focus on triangular transport maps, transformation random forests, and copulas. Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions. Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used. Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions. Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.

\section{Takeaways and roadmap}

A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail. Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests. Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results. Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.

>>>>

Multivariate density estimation is a cornerstone of modern statistics and machine learning. It supplies the machinery for likelihood-based inference, simulation of generative mechanisms, and uncertainty quantification in tabular domains. Practitioners, however, routinely face a delicate trade-off between three goals: expressive power, computational feasibility, and interpretability. Three complementary families of estimators have emerged around this dilemma:
\begin{enumerate}
  \item \textbf{Normalizing flows} build expressive change-of-variables maps and have become central in deep generative modeling \citep{rezende2015variational,papamakarios2017masked,dinh2017real,durkan2019neural,kingma2018glow,papamakarios2021normalizing}.
  \item \textbf{Transformation models and forests} model conditional distributions through parametric transformation structures coupled with ensemble methods \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation}.
  \item \textbf{Copulas} separate marginals from dependence, yielding interpretable multivariate association models \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence,nagler2017kdecopula}.
\end{enumerate}

This thesis situates all three families inside a common \textbf{transport framework} that makes them directly comparable \citep{rosenblatt1952remarks,knothe1957contributions,bogachev2005triangular,ramgraber2025friendly}. The framework operates in standardized coordinates $u$ computed solely from training-split statistics to avoid leakage and keep objectives aligned across models. Within this evaluation framework each estimator couples $u$ to a reference distribution---Gaussian for triangular maps and TRTF, $\mathrm{Unif}([0,1]^K)$ for copulas---and recovers densities via exact change of variables. Section~\ref{sec:ch1-frame} previews the core pieces and Chapter~\ref{ch:background} develops the mathematics in detail.

Separable triangular transport maps play a central role throughout the thesis. Their lower-triangular Jacobian delivers a log-determinant in $\mathcal{O}(K)$ operations, back-substitution yields an explicit inverse for sampling, and the triangular structure produces exact conditional densities in a fixed order. Fully triangular maps can require $\mathcal{O}(K^2)$ work per evaluation, whereas separable variants reduce both forward and inverse passes to $\mathcal{O}(K)$, making them attractive for systematic comparison while preserving interpretable marginals and dependence mechanisms.

\medskip
\noindent\textit{Notation.} We write $\varphi$ and $\Phi$ for the density and cumulative distribution function of a univariate standard normal random variable, and $\eta$ for the $K$-variate standard normal density. All models are evaluated in standardized coordinates $u$ and reported on the original scale $x$ through a diagonal affine correction so that log densities remain comparable. Unless stated otherwise, log values are expressed in natural units (nats).

\section{Thesis and Problem Statement}\label{sec:ch1-problem}

The goal of this thesis is to examine multivariate density estimation for tabular data within a unified transport-based evaluation protocol. We focus on three estimators: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and nonparametric copula models.

Let $u = D^{-1}(x - \mu)$ with $D = \mathrm{diag}(\sigma_1,\ldots,\sigma_K)$ determined from the training split. If $T: \mathbb{R}^K \to \mathbb{R}^K$ is lower triangular, $z = T(u)$, and $\eta$ denotes the chosen reference density, then the change-of-variables formula gives
\begin{equation}
  \log p_X(x) = \log \eta\big(T(u)\big) + \sum_{k=1}^{K} \log \frac{\partial T_k}{\partial u_k}(u_{1:k}) - \sum_{k=1}^{K} \log \sigma_k,
  \label{eq:ch1-change-of-variables}
\end{equation}
so reporting on the $x$-scale differs from reporting on the $u$-scale by the constant $-\sum_k \log \sigma_k$ per data point. The Rosenblatt and Knothe rearrangements guarantee the existence of such triangular transports for any coordinate ordering \citep{rosenblatt1952remarks,knothe1957contributions}.

The copula baselines considered here are nonparametric kernel copulas \citep{nagler2017kdecopula}. We fit smooth univariate marginals with \texttt{kde1d}, form pseudo-observations using mid-rank empirical CDFs during training to avoid double smoothing, and evaluate joint densities through
\[
  p_X(x) = c\big(F_1(x_1),\ldots,F_K(x_K)\big)\, \prod_{k=1}^{K} f_k(x_k),
\]
where $c$ is the copula density and $F_k = \int f_k$ the smoothed marginal CDFs. No Gaussian copula enters the study; the emphasis remains on flexible, interpretable diagnostics in low dimensions.

Separable triangular maps decompose each coordinate into a context-dependent shift plus a one-dimensional monotone transform. TRTF occupies an intermediate position: under strictly increasing conditional CDFs and the standard forest aggregation after monotone smoothing it reproduces the same separable triangular likelihood via the probability integral transform. The shared protocol described above keeps preprocessing, likelihood evaluation, calibration diagnostics, and compute measurements aligned across all estimators.

\section{The Transport Frame on One Page}\label{sec:ch1-frame}

The key ingredients of the transport framework are:
\begin{itemize}
  \item \textbf{Standardization.} Coordinates $u = D^{-1}(x-\mu)$ use training-split statistics only; all logs are in nats.
  \item \textbf{Change of variables.} Densities follow from $z = T(u)$ through Equation~\eqref{eq:ch1-change-of-variables}.
  \item \textbf{Triangular log determinant.} Lower-triangular Jacobians imply $\log \det J_T(u) = \sum_k \log \partial_{u_k} T_k(u_{1:k})$, giving $\mathcal{O}(K)$ cost.
  \item \textbf{Affine correction.} Reporting on $x$ applies the constant correction $-\sum_k \log \sigma_k$ per data point.
  \item \textbf{Computational profile.} Fully triangular maps can incur $\mathcal{O}(K^2)$ evaluation and inversion, whereas separable maps reduce both passes to $\mathcal{O}(K)$ while keeping the log determinant linear in $K$.
\end{itemize}

Separable triangular transports implement each coordinate as the sum of a context shift and a monotone one-dimensional basis, maintaining determinant stability and consistent conditional behaviour. TRTF achieves the same likelihood under the conditions noted above, while nonparametric copulas operate on the unit hypercube with explicit marginals. Standardization $u = T_{\mathrm{std}}(x)$ feeds either the triangular transport branch (TTM variants, TRTF) or the copula branch, and both return log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.

\section{Contributions and Research Questions}\label{sec:ch1-contributions}

The thesis contributes along three axes that move from theoretical unification to empirical benchmarking and practical interpretation:
\begin{enumerate}
  \item \textbf{Unified likelihood framework.} We formalise a shared transport perspective that places separable triangular maps, TRTF, and nonparametric copulas inside one mathematical scheme. The equivalence between TRTF and separable triangular transports holds under explicit regularity assumptions---strictly monotone conditional CDFs and the forest aggregation after smoothing.

    \begin{quote}
    \textbf{Proposition (informal).} Under axis-parallel partitions, strictly monotone conditional CDFs, and the forest aggregation after smoothing, the TRTF likelihood coincides with that of a separable triangular transport in standardized coordinates.
    \end{quote}

  \item \textbf{Benchmarking under a common protocol.} We design a standardized evaluation pipeline covering synthetic generators and real tabular benchmarks. TTM-Sep, its marginal variant, and TRTF are evaluated systematically; nonparametric kernel copulas act as diagnostic comparators when $K \le 3$. The pipeline produces three categories of evidence: (i) average test log likelihoods, (ii) conditional diagnostics based on probability integral transforms, and (iii) compute measurements for training effort and per-sample evaluation time. Together they quantify the trade-offs among accuracy, calibration, and efficiency.

  \item \textbf{Operational insights.} The theoretical and empirical findings yield guidance for practitioners. We analyse sensitivity to ordering, the practical limits of separability, and when nonparametric copulas remain informative as diagnostic references. Chapter~\ref{ch:conclusion} synthesizes these insights and outlines extensions toward richer parameterisations.
\end{enumerate}

Two research questions organise the empirical study:
\begin{itemize}
  \item \textbf{Synthetic data.} How do TRTF, TTM-Sep, and nonparametric copulas compare on controlled generators? Evidence comprises likelihood comparisons, conditional calibration metrics, and timing measurements.
  \item \textbf{Benchmark datasets.} How close do TRTF results come to published normalizing flow baselines under identical preprocessing? Side-by-side likelihood comparisons, interpreted through the constraints of separability and compute profiles, answer this question.
\end{itemize}

\begin{table}[t]
  \centering
  \caption{Model abbreviations used throughout the thesis.}
  \label{tab:model-abbrev}
  \begin{tabular}{ll}
    \hline
    Label & Meaning \\
    \hline
    TTM-Marg  & Marginal triangular transport (per-dimension; no context) \\
    TTM-Sep   & Separable triangular transport (context shift $+$ monotone 1D transform) \\
    TRTF      & Transformation Random Forests (axis-parallel splits) \\
    True-Marg & Oracle marginal density \\
    True-Joint& Oracle conditional joint density \\
    Copula    & Nonparametric kernel copula baseline (smoothed marginals; no Gaussian copula) \\
    \hline
  \end{tabular}
\end{table}

Addressing these questions bridges rigorous transport-based theory with empirical validation, clarifying when simple triangular models suffice and when more expressive families become necessary.
